F\# Data: \textnormal{Making structured data first-class citizens}

### 2015-02-27 icfp-2015/paper.tex
structured format/data vs structural value/type

% TODO: The model here does not actually use type erasure!!
% TODO: Maybe get rid of \mu_opt for records everywhere
% TODO: Disambigurate 's' as structural value and 's' as string
% TODO: Explain why we can only provide simple classes (!)
% TODO: Replace \nu_opt with \nu_? or something else that does not look shit
% TODO: Cleanup "strucutred" vs "structural"
% TODO: Insert a few more mentions of why "s1 + s2 <: s1 + s2 + s3" is valid (cannot do complete pattern match)

@@

To provide a basic intuition, the subtyping relation between structured types is illustrated in
Figure~\ref{fig:subtyping-diagram}. We split the diagram in two parts. The upper part shows 
non-nullable types (with records and primitive types). The lower part shows nullable types with 
\kvd{null}, collections and optional values. We omit links between the two part, but any type
$\hat{\sigma}$ is a subtype of $\hat{\sigma}~\kvd{option}$ (in the diagram, we abbreviate 
$\sigma~\kvd{option}$ as $\sigma?$).

@@

As mentioned earlier, the structured type inference relies on a finding common supertype. However,
the partially ordered set of types does not have a unique greatest lower bound. For example, record types
$\{\ident{a}:\ident{int}\}$ and $\{\ident{b}:\ident{bool}\}$ have common supertypes
$\{\ident{?a}:\ident{int}, \ident{?b}:\ident{bool}\}$ and $\{\ident{a}:\ident{int}\} + \{\ident{b}:\ident{bool}\}$
which are unrelated. Our inference algorithm always prefers records over unions (Theorem~\ref{thm:no-unions}). 
The definition of the \emph{common supertype} relation uses a number of auxiliary definitions 
that are explained in the discussion that follows.

@@

In this section, we build the necessary theoretical framework for proving the relativised type safety 
property of F\# Data type providers in Section~\ref{sec:safety}. We start by discussing the runtime 
representation of structured values and runtime conversions (Section~\ref{sec:formal-convert}). Then 
we embed these into a formal model of an F\# subset that is necessary for discussing type providers 
(Section~\ref{sec:formal-ff}) and we describe how F\# Data type providers turn an inferred 
structural type into actual F\# code (Section~\ref{sec:formal-tp}).

@@

In the model presented here, we represent JSON, XML and CSV documents using the same 
\emph{structural value}\footnote{Here, we diverge from the actual implementation in F\# Data which 
uses a different implementation for each format (reusing existing libraries).}. Structural values are 
first-order and can be one of the primitive values (Boolean, string, integer, decimal and float), a 
missing value (\kvd{null}), a collection and a record (named or unnamed):
%
\begin{equation*}
\begin{array}{lcl}
 s  &\narrow{=}& i \lsep d \lsep f \lsep t \lsep \kvd{true} \lsep \kvd{false} \lsep \kvd{null} \\[0.1em]
    &\narrow{|}& [s_1; \ldots; s_n] \\[0.1em]
    &\narrow{|}& \{ \nu_1 \mapsto s_1, \ldots, \nu_n \mapsto s_n \}\\[0.1em]
    &\narrow{|}& \nu~\{ \nu_1 \mapsto s_1, \ldots, \nu_n \mapsto s_n \} \\
\end{array}
\end{equation*}

@@

As suggested by subtyping, our system permits certain runtime conversions (\emph{e.g.}~a numerical 
value \num{1} can be treated as \kvd{true}). These are captured by the following operations:
%
\begin{equation*}
\begin{array}{lcl}
 op &\narrow{=}& \ident{asInt}(s) \lsep \ident{asDec}(s) \lsep \ident{asFloat}(s) \lsep \ident{asBool}(s)  \\
    &\narrow{|}& \ident{asStr}(s) \lsep \ident{getChildren}(s) \lsep \ident{getField}(\nu_{\ident{opt}}, \nu, s)\\
    &\narrow{|}& \ident{isNull}(s) \lsep \ident{isTag}(t, s)
\end{array}
\end{equation*}
%
The reduction rules for these operations are defined in Figure~\ref{fig:op-conversions}. It is important
to note that the relativised property we prove about F\# Data specifies \emph{sufficient}, but
not \emph{necessary} conditions. That is, a type provided based on a sample is guaranteed to work 
correctly on certain inputs, but may also handle several additional inputs.

@@

\noindent
In the above definition, $\tau$ denotes types including a type of all structured values $s$ called
\ident{StructVal}. A class definition $L$ consists of a constructor and zero or more members. Values $v$ include 
previously defined structural values $s$ and values for the option and list type; finally expressions $e$ include 
previously defined operations $op$, class construction, member access, conditionals and expressions for working 
with option values and lists. We include \ident{List.map} as a special construct to avoid making the language too complex.

@@

\paragraph{Type checking.} 
The type checking rules in Figure~\ref{fig:ff-typecheck} are written using a judgement
$L; \Gamma \vdash e : \tau$ where the context also contains a set of class declarations $L$.
The rules demonstrate the key difference between our language and standard ML or Featherweight Java:
%
\begin{itemize}[noitemsep]
\item[--] All structured values $s$ have a type \ident{StructVal}. Some of those have other types
  (primitive integers, decimals, floats, strings and Booleans as illustrated by the rule for $n$).
  For other values, \ident{StructVal} is the only type -- this includes records and \kvd{null}.
\item[--] A list containing other structural values $[s_1; \ldots; s_n]$ has a type \ident{StructVal}.  
  but can also have the $\ident{list}\langl\tau\rangl$ type. Conversely, lists that contain
  non-structural values like objects or options are not of type \ident{StructVal}.
\item[--] Primitive operations $op$ are treated as functions, accepting \ident{StructVal} and producing
  an appropriate type as the result.
\item[--] The two rules for checking class construction and member access are similar to corresponding
  rules of Featherweight Java.  
\end{itemize}
%
An important part of Featherweight Java that is omitted here is the checking of type declarations
(ensuring the bodies of members are well-typed). We omit this, because it is not needed for the
relativized safety theorem, which considers only classes generated by our type provider (which 
are correct by construction).

@@

So far, we defined the type inference algorithm for structured data formats which produces a structured
type $\sigma$ based on one or more sample documents (Section~\ref{sec:inference}). In Section~\ref{sec:formal-ff},
we defined a simplified model of F\# language and evaluation, which also included conversions that
model the F\# Data library runtime (Section~\ref{sec:formal-tp}). In this section, we define how
the type providers work, which links the two aspects discussed so far.

@@

\paragraph{Type provider mapping.}
When generating types, the type provider produces an F\# type $\tau$, expression that wraps a structured 
document value (of type \ident{StructVal}) as a value of type $\tau$ and a collection of class definitions. 
We express it using the following mapping:
%
\begin{equation*}
\sem{-}_e : \sigma \rightarrow \tau \times e' \times L
\end{equation*}
%
The semantics is parameterized by an expression $e$, which represents code to obtain a structured value that is
being wrapped. Then, given an inferred structured type $\sigma$, the type provider produces an F\# type $\tau$,
an expression $e'$ which constructs a value of type $\tau$ using $e$ and also a set of class definitions $L$.

@@

Handling of records is more interesting. We generate a new class $C$ that takes a structured value as constructor
parameter. For each record field, we generate a new member with the same name as the field\footnote{The actual
F\# Data implementation also capitalizes the names.} The body of the member calls \ident{getField} and then
passes this expression to $\sem{\sigma_i}$ which adds additional wrapping that maps the field (structured value
of type $\sigma_i$) into an F\# type $\tau_i$. The returned expression creates a new instance of $C$ and
the mapping returns the class $C$ together with all recursively generated definitions.

@@

\paragraph{Example 1.}
To illustrate how the type provision mechanism works, we now consider two simple examples. First, assume 
that the inferred type is a record with two fields (one optional) such as 
$\ident{Person}~\{~\ident{Age}:\ident{int}~\kvd{option}, \ident{Name}:\ident{string}~ \}$. Applying the
rules from Figure~\ref{fig:tp-generation} produces the following class:
%
\begin{equation*}
\begin{array}{l}
 \kvd{type}~\ident{Person}(\ident{v}:\ident{StructVal})~= \\[0.1em]
 \quad \kvd{member}~\ident{Age}~:~\ident{option}\langl\ident{int}\rangl~= \\[0.1em]
 \qquad \kvd{if}~\ident{isNull}~(\ident{getField}(\ident{Person},\ident{Age}, \ident{v}))~\kvd{then}~\ident{None} \\[0.1em]
 \qquad \kvd{else}~\ident{Some}(\ident{asInt}(\ident{getField}(\ident{Person},\ident{Age}, \ident{v}))) \\[0.1em]
 \quad \kvd{member}~\ident{Name}~:~\ident{string}~= \\[0.1em]
 \qquad \ident{asStr}(\ident{getField}(\ident{Person},\ident{Name}, \ident{v}))
\end{array}
\end{equation*}
%
The body of the \ident{Age} member is produced by the case for optional types applied to an expression
$\ident{getField}(\ident{Person},\ident{Age},v)$. If the returned field is not \kvd{null}, then the member
calls \ident{asInt} (produced by the case for primitive types) and wraps the result in the \ident{Some}
constructor. Note that \ident{getField} is defined even when the field does not exist, but returns \ident{null}.
This lets us treat missing fields as optional fields. The \ident{Name} member is similar, but does not 
perform any checks. 
For completeness, a type corresponding to the record is \ident{Person} and given a 
structural document $s$, we create a \ident{Person} value by calling $\kvd{new}~\ident{Person}(s)$.

@@

\subsection{Parsing structured values}
\label{sec:impl-parsing}

@@

The work that is most similar to F\# Data is the XML and SQL integration in C$\omega$ \cite{comega-xs}.
It extends a C\# like object-oriented language with types similar to our structured types 
(including nullable types, choices with subtyping and heterogeneous collections with multiplicities).
However, C$\omega$ does not infer the types from samples and extends the type system of the host
language (rather than using a general purpose embedding mechanism).

@@

The PADS project \cite{pads-dsl,pads-ml} tackles a more general problem of handling \emph{any} data format.
The schema definitions in PADS are similar to our structural type. The structure inference for LearnPADS
\cite{pads-learn} infers the data format from a flat input stream. A type provider for PADS could follow
many of the patterns we explore in this paper, but formally specifying the safety property would be
challenging.

>>>>

% TODO: The model here does not actually use type erasure!!
% TODO: Maybe get rid of \mu_opt for records everywhere
% TODO: Explain why we can only provide simple classes (!)
% TODO: Replace \nu_opt with \nu_? or something else that does not look shit
% TODO: Insert a few more mentions of why "s1 + s2 <: s1 + s2 + s3" is valid (cannot do complete pattern match)

@@

To provide a basic intuition, the subtyping relation between structural types is illustrated in
Figure~\ref{fig:subtyping-diagram}. We split the diagram in two parts. The upper part shows 
non-nullable types (with records and primitive types). The lower part shows nullable types with 
\kvd{null}, collections and optional values. We omit links between the two part, but any type
$\hat{\sigma}$ is a subtype of $\hat{\sigma}~\kvd{option}$ (in the diagram, we abbreviate 
$\sigma~\kvd{option}$ as $\sigma?$).

@@

As mentioned earlier, the structural type inference relies on a finding common supertype. However,
the partially ordered set of types does not have a unique greatest lower bound. For example, record types
$\{\ident{a}:\ident{int}\}$ and $\{\ident{b}:\ident{bool}\}$ have common supertypes
$\{\ident{?a}:\ident{int}, \ident{?b}:\ident{bool}\}$ and $\{\ident{a}:\ident{int}\} + \{\ident{b}:\ident{bool}\}$
which are unrelated. Our inference algorithm always prefers records over unions (Theorem~\ref{thm:no-unions}). 
The definition of the \emph{common supertype} relation uses a number of auxiliary definitions 
that are explained in the discussion that follows.

@@

In this section, we build the necessary theoretical framework for proving the relativised type safety 
property of F\# Data type providers in Section~\ref{sec:safety}. We start by discussing the runtime 
representation of structural values and runtime conversions (Section~\ref{sec:formal-convert}). Then 
we embed these into a formal model of an F\# subset that is necessary for discussing type providers 
(Section~\ref{sec:formal-ff}) and we describe how F\# Data type providers turn an inferred 
structural type into actual F\# code (Section~\ref{sec:formal-tp}).

@@

In the model presented here, we represent JSON, XML and CSV documents using the same 
\emph{structural value}\footnote{Here, we diverge from the actual implementation in F\# Data which 
uses a different implementation for each format (reusing existing libraries).}. Structural values are 
first-order and can be one of the primitive values (Boolean \kvd{true} or \kvd{false}, string $t$, 
integer $i$, decimal $d$ and floating-point number $f$), a missing value (\kvd{null}), a collection 
and a record (named or unnamed):
%
\begin{equation*}
\begin{array}{lcl}
 s  &\narrow{=}& i \lsep d \lsep f \lsep t \lsep \kvd{true} \lsep \kvd{false} \lsep \kvd{null} \\[0.1em]
    &\narrow{|}& [s_1; \ldots; s_n] \\[0.1em]
    &\narrow{|}& \{ \nu_1 \mapsto s_1, \ldots, \nu_n \mapsto s_n \}\\[0.1em]
    &\narrow{|}& \nu~\{ \nu_1 \mapsto s_1, \ldots, \nu_n \mapsto s_n \} \\
\end{array}
\end{equation*}

@@

As suggested by subtyping, our system permits certain runtime conversions (\emph{e.g.}~a numerical 
value \num{1} can be treated as \kvd{true}). These are captured by the following operations:
%
\begin{equation*}
\begin{array}{lcl}
 op &\narrow{=}& \ident{asInt}(s) \lsep \ident{asDec}(s) \lsep \ident{asFloat}(s) \lsep \ident{asBool}(s)  \\
    &\narrow{|}& \ident{asStr}(s) \lsep \ident{getChildren}(s) \lsep \ident{getField}(\nu_{\ident{opt}}, \nu, s)\\
    &\narrow{|}& \ident{isNull}(s) \lsep \ident{isTag}(\ident{tag}, s)
\end{array}
\end{equation*}
%
The reduction rules for these operations are defined in Figure~\ref{fig:op-conversions}. It is important
to note that the relativised property we prove about F\# Data specifies \emph{sufficient}, but
not \emph{necessary} conditions. That is, a type provided based on a sample is guaranteed to work 
correctly on certain inputs, but may also handle several additional inputs.

@@

\noindent
In the above definition, $\tau$ denotes types including a type of all structural values $s$ called
\ident{StructVal}. A class definition $L$ consists of a constructor and zero or more members. Values $v$ include 
previously defined structural values $s$ and values for the option and list type; finally expressions $e$ include 
previously defined operations $op$, class construction, member access, conditionals and expressions for working 
with option values and lists. We include \ident{List.map} as a special construct to avoid making the language too complex.

@@

\paragraph{Type checking.} 
The type checking rules in Figure~\ref{fig:ff-typecheck} are written using a judgement
$L; \Gamma \vdash e : \tau$ where the context also contains a set of class declarations $L$.
The rules demonstrate the key difference between our language and standard ML or Featherweight Java:
%
\begin{itemize}[noitemsep]
\item[--] All structural values $s$ have a type \ident{StructVal}. Some of those have other types
  (primitive integers, decimals, floats, strings and Booleans as illustrated by the rule for $n$).
  For other values, \ident{StructVal} is the only type -- this includes records and \kvd{null}.
\item[--] A list containing other structural values $[s_1; \ldots; s_n]$ has a type \ident{StructVal}.  
  but can also have the $\ident{list}\langl\tau\rangl$ type. Conversely, lists that contain
  non-structural values like objects or options are not of type \ident{StructVal}.
\item[--] Primitive operations $op$ are treated as functions, accepting \ident{StructVal} and producing
  an appropriate type as the result.
\item[--] The two rules for checking class construction and member access are similar to corresponding
  rules of Featherweight Java.  
\end{itemize}
%
An important part of Featherweight Java that is omitted here is the checking of type declarations
(ensuring the bodies of members are well-typed). We omit this, because it is not needed for the
relativized safety theorem, which considers only classes generated by our type provider (which 
are correct by construction).

@@

So far, we defined the type inference algorithm for structured data formats which produces a structural
type $\sigma$ based on one or more sample documents (Section~\ref{sec:inference}). In Section~\ref{sec:formal-ff},
we defined a simplified model of F\# language and evaluation, which also included conversions that
model the F\# Data library runtime (Section~\ref{sec:formal-tp}). In this section, we define how
the type providers work, which links the two aspects discussed so far.

@@

\paragraph{Type provider mapping.}
When generating types, the type provider produces an F\# type $\tau$, expression that wraps a structural 
document value (of type \ident{StructVal}) as a value of type $\tau$ and a collection of class definitions. 
We express it using the following mapping:
%
\begin{equation*}
\sem{-}_e : \sigma \rightarrow \tau \times e' \times L
\end{equation*}
%
The semantics is parameterized by an expression $e$, which represents code to obtain a structural value that is
being wrapped. Then, given an inferred structural type $\sigma$, the type provider produces an F\# type $\tau$,
an expression $e'$ which constructs a value of type $\tau$ using $e$ and also a set of class definitions $L$.

@@

Handling of records is more interesting. We generate a new class $C$ that takes a structural value as constructor
parameter. For each record field, we generate a new member with the same name as the field\footnote{The actual
F\# Data implementation also capitalizes the names.} The body of the member calls \ident{getField} and then
passes this expression to $\sem{\sigma_i}$ which adds additional wrapping that maps the field (structural value
of type $\sigma_i$) into an F\# type $\tau_i$. The returned expression creates a new instance of $C$ and
the mapping returns the class $C$ together with all recursively generated definitions.

@@

\paragraph{Example 1.}
To illustrate how the type provision mechanism works, we now consider two simple examples. First, assume 
that the inferred type is a record with two fields (one optional) such as 
$\ident{Person}~\{~\ident{Age}:\ident{int}~\kvd{option}, \ident{Name}:\ident{string}~ \}$. Applying the
rules from Figure~\ref{fig:tp-generation} produces the following class:
%
\begin{equation*}
\begin{array}{l}
 \kvd{type}~\ident{Person}(\ident{v}:\ident{StructVal})~= \\[0.1em]
 \quad \kvd{member}~\ident{Age}~:~\ident{option}\langl\ident{int}\rangl~= \\[0.1em]
 \qquad \kvd{if}~\ident{isNull}~(\ident{getField}(\ident{Person},\ident{Age}, \ident{v}))~\kvd{then}~\ident{None} \\[0.1em]
 \qquad \kvd{else}~\ident{Some}(\ident{asInt}(\ident{getField}(\ident{Person},\ident{Age}, \ident{v}))) \\[0.1em]
 \quad \kvd{member}~\ident{Name}~:~\ident{string}~= \\[0.1em]
 \qquad \ident{asStr}(\ident{getField}(\ident{Person},\ident{Name}, \ident{v}))
\end{array}
\end{equation*}
%
The body of the \ident{Age} member is produced by the case for optional types applied to an expression
$\ident{getField}(\ident{Person},\ident{Age},v)$. If the returned field is not \kvd{null}, then the member
calls \ident{asInt} (produced by the case for primitive types) and wraps the result in the \ident{Some}
constructor. Note that \ident{getField} is defined even when the field does not exist, but returns \ident{null}.
This lets us treat missing fields as optional fields. The \ident{Name} member is similar, but does not 
perform any checks. 
For completeness, a type corresponding to the record is \ident{Person} and given a 
structural value $s$, we create a \ident{Person} value by calling $\kvd{new}~\ident{Person}(s)$.

@@

\subsection{Parsing structured data}
\label{sec:impl-parsing}

@@

The work that is most similar to F\# Data is the XML and SQL integration in C$\omega$ \cite{comega-xs}.
It extends a C\# like object-oriented language with types similar to our structural types 
(including nullable types, choices with subtyping and heterogeneous collections with multiplicities).
However, C$\omega$ does not infer the types from samples and extends the type system of the host
language (rather than using a general purpose embedding mechanism).

@@

The PADS project \cite{pads-dsl,pads-ml} tackles a more general problem of handling \emph{any} data format.
The schema definitions in PADS are similar to our structural type. The structure inference for LearnPADS
\cite{pads-learn} infers the data format from a flat input stream. A PADS type provider could follow
many of the patterns we explore in this paper, but formally specifying the safety property would be
challenging.
